Clustering analysis and statistics

### 2019-05-05 notebooks/Clustering analysis.Rmd
removed some figures and changed headings

```{r libraries, include = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggtree)
library(cluster)
library(Biostrings)
library(ape)
library(tibble)
library(cowplot)
library(knitr)
library(distanceR)
```

@@

# Total dendrogram of all isolates
Genetic distances were calculated from the allele data from the cgMLST report, and a dendrogram was drawn from these distances. The annotations are based on data from the ARIBA resistance gene analysis.
```{r, fig.height=20, fig.width=20, message=FALSE, warning=FALSE}
cgMLST_data <- read.table("../data/chewbbaca/complete_results/clean_cgMLST_data.txt",
                   sep = "\t",
                   header = TRUE,
                   colClasses = "factor")

total_tree <- as.phylo(hclust(daisy(cgMLST_data, metric = "gower"), method = "average"))

tree_metadata <- read.table("../data/chewbbaca/complete_results/tree_metadata.txt",
                            sep = "\t",
                            header = TRUE)

tree_heatmap_data <- read.table("../data/chewbbaca/complete_results/tree_heatmap_data.txt",
                            sep = "\t",
                            header = TRUE) %>%
  select(-gadX)

names(tree_heatmap_data) <- c("GyrA","ParC","ParE","RpoB","MarA","MarR","SoxR","qepA4","qnr")

palette <- c("Broiler" = "#4575b4",
             "Pig" = "#74add1",
             "Red fox" = "#f46d43",
             "Wild bird" = "#fdae61")

palette2 <- c("1-A" = "#fc8d59",
              "1-I" = "#80b1d3",
              "0" = "grey95")

total_tree_annotated <- ggtree(total_tree,
                               layout = "circular",
                               color = "#808080",
                               size = 0.5) %<+% tree_metadata +
                               geom_tippoint(aes(color = species),
                               size = 3) +
                               geom_tiplab2(aes(label = ST),
                               offset = 0.01,
                               size = 4) +
  scale_color_manual(values = palette)

total_tree_opened <- rotate_tree(open_tree(total_tree_annotated, 8), 93)

gheatmap(total_tree_opened,
         tree_heatmap_data,
         offset = 0.04,
         width = 0.3,
         colnames_offset_y = 2.5,
         colnames_position = "top",
         font.size = 5) +
  scale_fill_manual(values = palette2,
                    labels = c("Gene/Mutation absent",
                               "Acquired gene present",
                               "Intrinsic gene mutated")) +
  theme(legend.position = c(0.9,0.9),
        legend.text = element_text(size = 20))
```

# Dendrograms per animal species
```{r, fig.height=25, fig.width=25, message=FALSE, warning=FALSE}

metadata_species <- read.table("../data/chewbbaca/complete_results/metadata_species_specific_trees.txt",
                               sep = "\t",
                               header = TRUE) %>%
  select(gyrA, parC, parE, PMQR) %>%
  dplyr::rename("GyrA" = gyrA,
         "ParC" = parC,
         "ParE" = parE)

palette3 <- c("0" = "grey95",
              # gyrA
              "S83L" = "#c6dbef",
              "S83A" = "#9ecae1",
              "D87N" = "#6baed6",
              "D87H" = "#4292c6",
              "D87Y" = "#2171b5",
              "D87G" = "#08519c",
              "S83L, D87N" = "#08306b",
              # parC
              "S80I" = "#a1d99b",
              "S80R" = "#74c476",
              "S57T" = "#41ab5d",
              "A56T, S80I" = "#238b45",
              "S80I, E84V" = "#005a32",
              # parE
              "D475E" = "#fdd0a2",
              "D463N" = "#fdae6b",
              "S458A" = "#fd8d3c",
              "L416F" = "#f16913",
              "H516Y" = "#d94801",
              "L488M, A512T" = "#8c2d04",
              # PMQR
              "qnrA1" = "#dadaeb",
              "qnrB19" = "#bcbddc",
              "qnrS1" = "#9e9ac8",
              "qnrS2" = "#807dba",
              "qnrS4" = "#6a51a3",
              "qepA4" = "#4a1486")

# Broiler tree

broiler_tree <- calc_tree("../data/chewbbaca/complete_results/cgMLST_broiler.tsv")

broiler_tree_annotated <- ggtree(broiler_tree,
                                 layout = "circular",
                                 color = "#808080",
                                 size = 0.5) %<+% tree_metadata +
  geom_tippoint(color = "#4575b4",
                size = 4) +
  geom_tiplab2(aes(label = ST),
               offset = 0.02,
               size = 6)

broiler_complete <- gheatmap(broiler_tree_annotated,
         metadata_species,
         offset = 0.1,
         width = 0.5,
         colnames = FALSE) +
  scale_fill_manual(values = palette3) +
  guides(fill = FALSE)

# Pig tree
pig_tree <- calc_tree("../data/chewbbaca/complete_results/cgMLST_pig.tsv")

pig_tree_annotated <- ggtree(pig_tree,
                                 layout = "circular",
                                 color = "#808080",
                                 size = 0.5) %<+% tree_metadata +
  geom_tippoint(color = "#74add1",
                size = 4) +
  geom_tiplab2(aes(label = ST),
               offset = 0.02,
               size = 5.5)

pig_complete <- gheatmap(pig_tree_annotated,
         metadata_species,
         offset = 0.09,
         width = 0.5,
         colnames = FALSE) +
  scale_fill_manual(values = palette3) +
  guides(fill = FALSE)

# Red fox tree
fox_tree <- calc_tree("../data/chewbbaca/complete_results/cgMLST_fox.tsv")

fox_tree_annotated <- ggtree(fox_tree,
                                 layout = "circular",
                                 color = "#808080",
                                 size = 0.5) %<+% tree_metadata +
  geom_tippoint(color = "#f46d43",
                size = 4) +
  geom_tiplab2(aes(label = ST),
               offset = 0.02,
               size = 5.5)

fox_complete <- gheatmap(fox_tree_annotated,
         metadata_species,
         offset = 0.09,
         width = 0.5,
         colnames = FALSE) +
  scale_fill_manual(values = palette3) +
  guides(fill = FALSE)

# Wild bird tree
bird_tree <- calc_tree("../data/chewbbaca/complete_results/cgMLST_bird.tsv")

bird_tree_annotated <- ggtree(bird_tree,
                                 layout = "circular",
                                 color = "#808080",
                                 size = 0.5) %<+% tree_metadata +
  geom_tippoint(color = "#fdae61",
                size = 4) +
  geom_tiplab2(aes(label = ST),
               offset = 0.02,
               size = 5.5)

bird_complete <- gheatmap(bird_tree_annotated,
         metadata_species,
         offset = 0.09,
         width = 0.5,
         colnames = FALSE) +
  scale_fill_manual(values = palette3) +
  guides(fill = FALSE)

# Total plot
tot_plot <- plot_grid(broiler_complete,
          pig_complete,
          fox_complete,
          bird_complete,
          nrow = 2,
          ncol = 2,
          align = "h",
          labels = c("Broiler","Pig","Red fox","Wild bird"))

include_graphics("../figures/species_dendrogram2.png")
```

@@

# Statistics
## Distribution of calculated distances per animal species
Below follows a plot of the distribution of the lower triangular distances for each animal species. A bimodal distribution is present in all four plots. The filtering values of < 0.1 and > 0.8 is used for downstream statistics to acertain whether the broiler isolates are more homogenous than the isolates from other species.

```{r}
# calculates distances from cgMLST data and returns
# the lower triangular of the matrix, with (diag = FALSE)
# or without (diag = TRUE) the diagonal
lower_tri_dist_calc <- function(data, diag = TRUE, data_frame = FALSE) {
  dist <- as.matrix(daisy(data, metric = "gower"))
  if (diag == TRUE) {
    dist[upper.tri(dist, diag = TRUE)] <- NA
  } else {
    dist[upper.tri(dist, diag = FALSE)] <- NA
  }
  dist_vec <- as.vector(as.matrix(dist))
  dist_compl <- dist_vec[!is.na(dist_vec)]

@@

## Distribution of iterated distances
Distribution of percentages of distances < 0.1 or > 0.8. The distribution is based on 1000 iterations of the distance data. Observed values are denoted as arrows for each animal species; dark blue = Broiler, light blue = Pig, red = Red fox, yellow = Wild bird. The mean value is denoted as a vertical black line.
```{r, warning=FALSE, message=FALSE, dpi=300}
# generate data frame from allele data
alleles <- cgMLST_allele_data %>%
  left_join(isolate_data[,c("id","species")], by = "id") %>%
  group_by(species) %>%
  mutate(n = 1:n(),
         new_id = case_when(species == "Broiler" ~ paste("BR", n, sep = "_"),
                            species == "Pig" ~ paste("PI", n, sep = "_"),
                            species == "Red fox" ~ paste("RF", n, sep = "_"),
                            species == "Wild bird" ~ paste("WB", n, sep = "_"))) %>%
  ungroup() %>%
  select(new_id, everything(), -c(species, id)) %>%
  mutate_all(funs(as.factor)) %>%
  column_to_rownames("new_id")

@@

>>>>

```{r libraries, include = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggtree)
library(cluster)
library(Biostrings)
library(ape)
library(tibble)
library(cowplot)
library(knitr)
library(vegan)
library(distanceR)
```

@@

@@

# Allele distance distribution
```{r}
# calculates distances from cgMLST data and returns
# the lower triangular of the matrix, with (diag = FALSE)
# or without (diag = TRUE) the diagonal
lower_tri_dist_calc <- function(data, diag = TRUE, data_frame = FALSE) {
  dist <- as.matrix(daisy(data, metric = "gower"))
  if (diag == TRUE) {
    dist[upper.tri(dist, diag = TRUE)] <- NA
  } else {
    dist[upper.tri(dist, diag = FALSE)] <- NA
  }
  dist_vec <- as.vector(as.matrix(dist))
  dist_compl <- dist_vec[!is.na(dist_vec)]

@@

# Distribution of iterated distances
Distribution of percentages of distances < 0.1 or > 0.8. The distribution is based on 1000 iterations of the distance data. Observed values are denoted as arrows for each animal species; dark blue = Broiler, light blue = Pig, red = Red fox, yellow = Wild bird. The mean value is denoted as a vertical black line.
```{r, warning=FALSE, message=FALSE, dpi=300}
# generate data frame from allele data
alleles <- cgMLST_allele_data %>%
  left_join(isolate_data[,c("id","species")], by = "id") %>%
  group_by(species) %>%
  mutate(n = 1:n(),
         new_id = case_when(species == "Broiler" ~ paste("BR", n, sep = "_"),
                            species == "Pig" ~ paste("PI", n, sep = "_"),
                            species == "Red fox" ~ paste("RF", n, sep = "_"),
                            species == "Wild bird" ~ paste("WB", n, sep = "_"))) %>%
  ungroup() %>%
  select(new_id, everything(), -c(species, id)) %>%
  mutate_all(funs(as.factor)) %>%
  column_to_rownames("new_id")

@@

# NMDS Analysis of resistance mechanisms
Non-metric multidimensional scaling (NMDS) was used to see if the distribution of quinolone resistance mechanisms from some sequence types were more homogenous than in other sequence types.
```{r}
res_mechanisms <- read.table("../data/MDS_data.txt",
                       sep = "\t",
                       header = TRUE,
                       stringsAsFactors = FALSE) %>%
  column_to_rownames("id")

bin_dist <- dist(res_mechanisms,
                 method = "binary")

nmds_df <- metaMDS(bin_dist,
                   trymax = 200,
                   trace = FALSE) # silence the output

print(paste0("Stress: ", round(nmds_df$stress, 3)))

stressplot(nmds_df)
